Relationships in the extremes and their influence on competition and coexistence

### 2022-04-21 paper/Paper.Rnw
made some progress on the Discussion

We produced and applied a new extension of MCT that 
quantifies the impact of ATAs on species coexistence and the maintenance of biodiversity. 
Building upon a recent, simulation-based approach to MCT \citep{Ellner_2016, Ellner_2019},
we decomposed storage effects into two new mechanisms, the contribution of ATAs to coexistence,
and the contribution of $EC$ correlation \emph{per se}.
Applications of our extended theory to the lottery model and to models based on a laboratory 
system of competing diatoms demonstrated that  
ATA contributions to storage effects and to species coexistence and the maintenance of biodiversity
can often be comparable in magnitude to the contributions of previously recognized mechanisms.
Our results highlight that common methods of correlation and covariance fail to include 
important information on the structure of relationships between variables. Our results add
coexistence and biodiversity maintenance to the list of ecological theories and phenomena on which 
ATAs have been shown to have an important impact \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of $EC$ \emph{covariance} to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019},
our results suggest that storage effects should be redefined to constitute the 
contribution of \emph{associations} between $E$ and $C$, including both ATA effects and
correlation \emph{per se}; the prior terminology, \emph{covariance}, is typically construed 
as reflecting the linear relationship between variables, ignoring complexities of the association,
such as ATAs, which we now know are important.

%ATA specialists
Contributions of ATAs to species coexistence open the possibility that
``ATA specialists'' may exist in some contexts. If a strong competitor is present across a
network of habitat patches and the GWR of a weaker competitor is only positive 
in a patch in the presence of 
ATAs between the temporal fluctuations of two environmental variables in that patch,
then the weaker competitor can only persist in patches for which ATAs are present,
and can therefore be regarded as an ``ATA specialist'' in the landscape. EC-associations
can often boil down to associations between two environmental variables, as was the case
for the lottery model we considered. Mechanisms of coexitence should correspond to
means by which niche differentiation is possible. Thinking about our new mechanism of 
coexistence in terms of possible ATA specialists helps highlight this connection. Likewise,
because we have described ATA exclusion as well as ATA rescue, it is possible to 
imagine a scenario in which a weaker competitor can only exist in habitat patches 
for which ATAs are absent, and is hence a specialist on scenarios with symmetric
tail associations of environmental variables (an ``STA-specialist''). Additional 
research will be needed to assess to what degree these possibilities are realized 
in nature.

%future work: study ATAs in environmental variables and how theyre changing - tie in with specialist
Our study highlights the importance of ATAs for ecological phenomena 
and specifically for coexistence and the maintenance of biodiversity. 
The occurence of ATAs among environmental variables themselves, 
their contribution to coexistence, and their changes through space and time
can and should be explored further. 
The coupling of ATA contribution magnitude with environmental variability, 
as shown in our results and figures (Fig. \ref{fig:fig3} - \ref{fig:fig6}), 
suggests that ATAs and their effects
may become more important to consider for coexistence and possibly other 
ecological phenomena as environmental conditions may evidently
become more variable (probably need refs).
As described in section \ref{section:introduction}, ATAs between environmental
variables may shift as the climate becomes more variable and extreme events become more 
expansive and extensive, causing associations of distinct temporal events to become 
more related in one extreme or tail, thus asymmetric. Additionally, 
if ATA specialist exist, their persistance may
become at stake as the mechanism driving their coexistence changes with the climate.
Thus it is of interest to further study how tail associations of environmental variables are changing and how this might 
alter competition, coexistence, and the maintenance of biodiversity generally.
Our investigation brings to attention the consequences of such
possible [bivariate] \emph{relationships} in the extremes; however, 
simpler changes in [the marginal/ univariate distributions of] extreme events themselves such as
intensity and frequency and their influence on coexistence also merit future work. 
Possible changes in tail association
patterns between environmental variables in nature can
influence the coexistence of species through the storage effect and not other mechanisms, but
extreme and changing climate is probable to affect other mechanisms in other ways (refs?).
%In the Discussion, I intend to talk about how possible changes in tail association 
%patterns of env vars in nature may influence coexistence of species. It won't influence the 
%other mechanisms, though changes in the extreme values of environmental variables, known
%to be occurring, might. 
%Should also mention this second point in the Discussion as something that merits further research.
%though might be worth mentioning in the Discussion that events are getting more extreme. We have studied how *relationships*
%in the extremes influence coexistence, but probably worth studying (or cite papers if it's been done) how changes
%in the commonness and extremeness of extreme events themselves influence coexistence, and for that you'd have
%to also look at the other mechanisms. May still need a sup mat laying out all the math.

%Although environmental fluctuations is helpful for coexistence and storage effect, unusual extreme fluctuations can be a hinderance on coexistence by introducing or changing ATAs between environmental variables. 
%onsistent with the mechanisms of biodiversity maintenance and specifically 
%luctuation-dependent mechanisms (cite chesson), higher values of environmental fluctuation 
%esulted in higher IGR’s. Both of our new terms, $\Delta_i^{[EC]}$ and $\Delta_i^{[E \nat C]}$ 
%lso increased in magnitude with increased environmental fluctuation. Increased environmental 
%luctuation is reflected in increased $\sigma$ in the lottery model and increased $a$ in the 
%ompetitive diatom model.   mean temp and period?

%This would be a whole potential 
%research program which presumably would start with a more detailed study of how tail associations of environmental
%variables may be changing under global change.

%\underline{E and vital rates}
We note that \emph{E}, in our investigation and in other investigations within the MCT 
framework, is not directly a measure of the environment, but is a measure of response to the 
environment, via a population's vital rates. Environment and environmental response may be related linearly
(like lottery model?); though real environmental variables are probable to 
actually be related to vital rates in a non-linear and often threshold-like way. 
Threshold relationships can produce tail association, as we showed elsewhere \citep{Walter_2022}. 
Changes in the means of true environmental variables may change 
where those distributions of environmental variables are compared to thresholds, 
which may change the nature of tail associations between the effects of 
those environmental variables. Thus, not only do the possible asymmteric association between 
environmental variables merit future study, but the relationships between populations'
vital rates should also be investigated for species coexistence and the maintenance of biodiversity.
%Discussion content: We should (in future) study tail associations between environmental variables, and how these might be changing. However, more than that, the variables E in the lottery model are not actual environmental variables but rather are the effects of environmental variables on the species of the model.

%\underline{the new effects are not "weak noise" effects}
%if sigma really small, ATA would not matter?
%go back to emails
%Maybe have a paragraph about how Chesson's weak noise and other assumptions make it impossible to detect this new effect
%via that framework. It's only the less-assumption-reliant simulation approach of Ellner et al that makes it possible to 
%study this new term in the decomp. Need to study a bit to substantiate this properly, or get Steve Ellner help.
We parse the storage effect, providing a tertiary decomposition to the simulation-based 
approach of MCT \citep{Ellner_2016, Ellner_2019}, 
a new approach which bypasses the mathematical assumptions and obstructions of 
standard MCT. Classical analytic derivations of MCT using second order Taylor expansion and 
small-variance approximations completely miss out
on any additions to coexistence beyond simple correlations and does not include our new ATA term, $\Delta^{[EC]}$, or variance \emph{per se}, $\Delta^{(E \shp C)}$ \citep{Chesson_1994, Chesson_2000}. 
In fact, storage effects in analytic MCT is computationally equal to $\Delta^{[E \nat C]}$, 
the contributions of correlation \emph{per se}. 
In newer, simulation based MCT, storage effect is 
the sum of the two new terms, $\Delta^{(EC)}=\Delta^{[E \nat C]}+\Delta^{[EC]}$, 
obscuring the effects of ATAs. Because standard MCT's assumption make it impossible to 
detect interactions outide of correlation \emph{per se}, simulation-based MCT is required to
detect ATA contributions, and our new extension to it clarifies and cleary separates and quantifies the 
relative importance of both ATAs and pure correlation in contribution to the storage effect
and subsequently species coexistence and the maintenance of biodiversity.

%\underline{multispecies models}
%Need a para about multi-species models - how would things change? One difference is there are more different
%things that might have asymmetric tail association patterns. See the section "General Theory" of Ellner et al 2019 - might
%just want write a short para saying it seems like it should be possible to generalize what we've done along those lines.

%Because asymmetric tail associations exist in many ecological variables \citep{Ghosh_2020_AER}, the more specified approach of considering unique drivers of coexistence rather than E and C and the ability to consider more than two drivers, could produce interesting results??

%The inclusion of multiple species introduces a multitude of relationships between species'
%environmentally dependent parameters that could possibly be asymmetrically associated. 
%The greater the number of species the greater the possibilities of ATAs being more important 
%to their coexistence.

The chances of ATAs being important for the coexistence of species increases as more variables are included in the system, 
either through the addition of more species or more drivers (beyond just E and C), or both. 
When multiple species are included in the system, the interaction between different species' E's and C's introduces more relationships 
to consider that may or may not be asymmetrically related and could increase the relative importance of ATA contribution to the species' coexistence. 
Likewise, coexistence between species can also depend on more than two drivers \citep{Ellner_2019}. 
The storage effect decomposition presented here is generalizable to coexistence decompositions of multiple variables.
Following the set up and notation of \cite{Ellner_2019}'s \emph{covariance decomposition}, 
ATA contributions and correlation \emph{per se} can be decomposed from the \emph{associations} between two variables
among many from the appropriate term with parenthesized superscripts containing the $\shp$ symbol which denotes and separates 
variables that are varying independently from the variables that are covarying whose relationship we are interested in decomposing. 
For example, the term $\Delta^{(XY \shp Z)}$ could be decomposed into $\Delta^{[XY \shp Z]} + \Delta^{[X \nat Y \shp Z]}$. 
As more terms and species are introduced into a system, more relationships between variables must be evaluated,
more of those relationships may be asymmetrically tail associated, and ATAs may contribute more to the maintenance of 
biodiversity in that system. 
%make three different biv dis that are standard normal copula and then pair with a shuffles third
%use to get rbar and subtract from rbar of XY#Z,

>>>>

%Summary-type paragraph of main results
We produced and applied a new extension of MCT that 
quantifies the impact of ATAs on species coexistence and the maintenance of biodiversity. 
Building upon a recent, simulation-based approach to MCT \citep{Ellner_2016, Ellner_2019},
we decomposed storage effects into two new mechanisms, the contribution of ATAs to coexistence,
and the contribution of $EC$ correlation \emph{per se}.
Applications of our extended theory to the lottery model and to models based on a laboratory 
system of competing diatoms demonstrated that  
ATA contributions to storage effects and to species coexistence and the maintenance of biodiversity
can often be comparable in magnitude to the contributions of previously recognized mechanisms.
Our results highlight that common methods of correlation and covariance fail to include 
important information on the structure of relationships between variables. Our results add
coexistence and biodiversity maintenance to the list of ecological theories and phenomena on which 
ATAs have been shown to have an important impact \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of $EC$ \emph{covariance} to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019},
our results suggest that storage effects should be redefined to constitute the 
contribution of \emph{associations} between $E$ and $C$, including both ATA effects and
correlation \emph{per se}; the prior terminology, \emph{covariance}, is typically construed 
as reflecting the linear relationship between variables, ignoring complexities of the association,
such as ATAs, which we now know are important.

%ATA specialists
Contributions of ATAs to species coexistence open the possibility that
``ATA specialists'' may exist in some contexts. If a strong competitor is present across a
network of habitat patches and the GWR of a weaker competitor is only positive 
in a patch in the presence of 
ATAs between the temporal fluctuations of two environmental variables in that patch,
then the weaker competitor can only persist in patches for which ATAs are present,
and can therefore be regarded as an ``ATA specialist'' in the landscape. EC-associations
can often boil down to associations between two environmental variables, as was the case
for the lottery model we considered. Mechanisms of coexitence should correspond to
means by which niche differentiation is possible. Thinking about our new mechanism of 
coexistence in terms of possible ATA specialists helps highlight this connection. Likewise,
because we have described ATA exclusion as well as ATA rescue, it is possible to 
imagine a scenario in which a weaker competitor can only exist in habitat patches 
for which ATAs are absent, and is hence a specialist on scenarios with symmetric
tail associations of environmental variables (an ``STA specialist''). Additional 
research will be needed to assess to what degree these possibilities are realized 
in nature.

%future work: study ATAs in environmental variables and how theyre changing - tie in with specialist
In the Introduction we explained a plausible means by which climate change may
alter the tail association properties of phenologically separated environmental variables
such as early- and late-spring rainfall if, as seems likely, meteological events which
are more extreme also tend to be longer lasting. The possibility 
that changes in the nature of environmental ATAs 
may result in the extirpation of ATA and STA specialists indicates that it may
be important to engage in further study of ATAs between environmental variables 
and how they differ across space and may be changing through time due to human
impacts on the climate. Extreme environmental events have been much studied, including
how the nature of such events may be changing with climate change, and how such changes 
may influence species coexistence (see citations in the Introduction). But
studies of changes in environmental ATAs would differ from such prior work because
ATAs pertain to relationships between variables in the extremes, a potential changes in those
relationships, to the best of our knowledge, have not yet been examined.

%E and vital rates
Whereas $E$ of the $EC$ associatons which produce storage effects 
is often conceptualized as ``environment'', in fact it is a 
species response to the environment. Because a directly measurable aspect of the 
environment and a species response to it may be nonlinearly related, it may be 
important, in future work, 
to also study the nature of tail associations between species vital rates.
In the lottery model, it was assumed that fecundity of species $i$ tracks the
environment directly, $B_i(t)=E_i(t)$. But vital rates such as fecundity are 
more likely related to actual environmental variables according to a nonlinear
function which may often be threshold-like or sigmoidal, and thus the $E_i=B_i$
of the lottery model is actually better conceptualized as a species response
than as the environment, directly.
Likewise, one of the choices of $E_i$ considered by \cite{Ellner_2016} for the
diatom model was $E_i=V_i$, the maximum population growth rate, clearly a response
to the environment rather than a direct measurement of the environment. 
It was demonstrated elsewhere that nonlinear transformations can produce or alter 
asymmetries of tail association \citep{Ghosh_2020_AER,Walter_2022}.
Thus the nature of ATAs between directly measured environmental variables may differ from the
nature of ATAs between the population quantities which actually matter for species coexistence.
Vital rates are often measured directly in demongraphic studies. To our knowledge, the 
nature of tail associations between vital rates has been studied seldom or never.

%Differences with the original, analytic version of the MCT
%
%***DAN: I am not so comfortable with this because I have insufficient 
%knowledge of the original MCT. I also do not want to devote the time
%to learning it right now because of other commitments. I've reworded, 
%trying not to change the meaning. Jasmin, how confident are you in 
%these statements? We will also want to try to get Steve Ellner to check 
%them.
Our approach to exploring the influence of ATAs on coexistence relies on the computational
extension of the MCT of \cite{Ellner_2019},
and our results help demonstrate strengths of that approach.
Storage effects as computed using the original, analytic appoach to MCT 
\citep{Chesson_1994,Chesson_2000} should equal our term $\Delta^{[E \nat C]}$, 
which we refer to as the contributions to coexistence of correlation \emph{per se}.
This is because the weak-noise/Taylor approximation approach used in the original 
MCT neglects the effects of ATAs when higher-order terms
in Taylor expansions are dropped. The term $\Delta^{(E \shp C)}$, which our theory 
inherits from \cite{Ellner_2016}, is also dropped. These observations do not diminish 
the major strengths of the 
original, analytic appproach, which are well known and also elaborated by \cite{Ellner_2019};
one such strength is the generality that analtic approaches provide.
It may be possible to extend the mathematics of the original MCT to consider ATAs.

%multispecies models and multifactor models
\cite{Ellner_2019} elaborated a general theory in which more and different varying factors than
$E$ and $C$ can be considered and used to produce a decomposition of GWR. For instance, if $X$, $Y$ and 
$Z$ are temporally varying factors, \cite{Ellner_2019} produced, for the $i^{\text{th}}$ species, the decomposition
\begin{align}
\text{GWR}_i &= \Delta_i^0+\Delta_i^X+\Delta_i^Y+\Delta_i^Z \label{eq:9}\\
&+\Delta_i^{(X \shp Y)} +\Delta_i^{(XY)} + \Delta_i^{(Y \shp Z)} + \Delta_i^{(YZ)} +\Delta_i^{(X \shp Z)}+\Delta_i^{(XZ)} \label{eq:10}\\
&+ \Delta_i^{(XYZ)}+\Delta_i^{(XY \shp Z)}+\Delta_i^{(X \shp YZ)}+\Delta_i^{(XZ \shp Y)}+\Delta_i^{(X \shp Y \shp Z)}.
\end{align}
The terms in (\ref{eq:9}) and (\ref{eq:10}) are analogous to those of (\ref{eq:previous_decomp}).
The new terms result from rendering one or more of $X$, $Y$, $Z$ independent of the others.
We simply comment that this general decomposition should also be extendable to consider the effects of
ATAs in the same way (\ref{eq:full_decomp}) extends (\ref{eq:previous_decomp}), though the complexity
of such a decomposition may present some difficulties.
For example, $\Delta^{(XY \shp Z)}$ can be decomposed as $\Delta^{[XY \shp Z]} + \Delta^{[X \nat Y \shp Z]}$.
It seems reasonable to suppose that ATAs will play an increasingly important role in systems of 
greater complexity because such systems have more pairs of temporally variable quantities that
may exhibit asymmetries of association. Our prior work shows that ATAs are common in 
ecological and environmental variables \citep{Ghosh_2020_AER}.

%***DAN: stopped here 2022 04 21
